---
title: "Exploratory Analysis"
author: "Addison McGhee"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = TRUE, message = F)
```


```{r, echo=F, include=F}
library(tidyverse)
library(pander)
library(ggrepel)
library(splines2)
library(splines)
library(ggthemes)
library(gam)
library(gridExtra)
library(foreign)
library(readr)
library(pander)
library(aod)
theme_update(plot.title = element_text(hjust = 0.5))
```

### Read in data and create factors
```{r}
dat = read_csv("heart.csv")
dataset_wfactors = dat %>% mutate(anaemia = factor(anaemia),
                             diabetes = factor(diabetes),
                             high_blood_pressure = factor(high_blood_pressure),
                             sex = factor(sex),
                             smoking = factor(smoking),
                             DEATH_EVENT = factor(DEATH_EVENT)
                             )

# Dataset without factors
dataset = dataset_wfactors %>% select(-c(anaemia, diabetes, high_blood_pressure, sex, smoking, DEATH_EVENT))
```

**As with any exploratory analysis, the first step is to determine which variable are categorical, and which are continuous. We see that there are 6 categorical variables in this dataset. They were converted to `factors` to ensure ease of use with $R's$ functions.**

&nbsp;

## Missing Values
```{r}
apply(dataset_wfactors, 2, FUN = function(x) sum(is.na(x)))
```

**The next step was to screen for missing values as many $R$ functions cannot handle NAs. In addition, missing values can be due to underlying issues with the treatment and/or data collection procedure. Fortunately, the data does not have any missingness.**

&nbsp;

## Summary Stats

### Continuous Correlation Matrix
```{r, echo=F}
cormat <- round(cor(dataset), 2)

library(reshape2)

# Get lower triangle of the correlation matrix
get_lower_tri<-function(cormat){
    cormat[upper.tri(cormat)] <- NA
    return(cormat)
}
  
  # Get upper triangle of the correlation matrix
get_upper_tri <- function(cormat){
    cormat[lower.tri(cormat)]<- NA
    return(cormat)
}
  
reorder_cormat <- function(cormat){
# Use correlation between variables as distance
dd <- as.dist((1-cormat)/2)
hc <- hclust(dd)
cormat <-cormat[hc$order, hc$order]
}


# Reorder the correlation matrix
cormat <- reorder_cormat(cormat)
upper_tri <- get_upper_tri(cormat)

# Melt the correlation matrix
melted_cormat <- melt(upper_tri, na.rm = TRUE)

# Create a ggheatmap
ggheatmap <- ggplot(melted_cormat, aes(Var2, Var1, fill = value))+
 geom_tile(color = "white")+
 scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
    name="Pearson\nCorrelation") +
  theme_minimal()+ # minimal theme
 theme(axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 12, hjust = 1))+
 coord_fixed()


ggheatmap + 
geom_text(aes(Var2, Var1, label = value), color = "black", size = 2) +
theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.grid.major = element_blank(),
  panel.border = element_blank(),
  panel.background = element_blank(),
  axis.ticks = element_blank(),
  legend.justification = c(1, 0),
  legend.position = c(0.6, 0.7),
  legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                title.position = "top", title.hjust = 0.5))
```

&nbsp;

**The above graphic indicates positive and negative associations between variables. There also is a clear clustering among certain variables that share a common direction of association. We notice moderate negative relationshps between time/age, time/serum_creatinine, and serum_sodium/serum_creatinine. Our next step will be to gain additional insight into the biological mechanisms that explain why certain variables associate with each other.**

&nbsp;

## Histograms

### Continuous Variables
```{r}
do = data.frame(dataset)
for(i in 1:ncol(do)) {
    h = hist(do[, i], main = names(do)[i], xlab = names(do)[i])
    text(h$mids,h$counts,labels=h$counts, adj=c(0.5, -0.5), cex = 0.6, col = "red")
}
```

&nbsp;

**The histograms showcase the different distributions present in the data. The variables of serum_creatinine, creatinine_phosphokinase, and age have a right-skew, while serum_sodium, platelets, and ejection_fraction are more normally distributed. The variable time is more uniformly distributed. Again, more domain knowledge will give us a better understanding of the typical distributions that these variables take. This will be useful in helping us gauge the generalizability of our results.**

&nbsp;

### Factors
```{r}
dataset_wfactors %>% select(c(anaemia, diabetes, high_blood_pressure, sex, smoking, DEATH_EVENT)) %>% apply(2, table) %>% pander
```

```{r}
dataset_wfactors %>% group_by(sex) %>% count(DEATH_EVENT) %>% pander
dataset_wfactors %>% group_by(smoking) %>% count(DEATH_EVENT) %>% pander
dataset_wfactors %>% group_by(anaemia) %>% count(DEATH_EVENT) %>% pander
```

**The above tables show a common ration of `2 to 1` for many of the factors. Smoking/non-Smoking is roughly 2:1, as is high/not-high blood pressure. This is also true when we look at the counts for Death-event for each sex; the ratio of dying to not dying is roughly `2 to 1` for both men and women.**


&nbsp;

## Regression Models

### Initial Logistic Screening (all variables)
```{r}
logistic_1 <- glm(data = dataset_wfactors,
               DEATH_EVENT ~ ., family = "binomial")


pander(summary(logistic_1))
```

&nbsp;

**The inital screening above provides evidence of which terms might be most useful in predicting the death_event outcome. In particular, it seems that age, ejection_fraction, serum_creatinine, and time are the most significant covariates in our model. Further steps will be to use other metrics (AIC, BIC, etc.) to perform model selection and assess our model.**

&nbsp;

### 95% CI for Odds-Ratios
```{r}
logistic_1 %>% confint %>% exp %>% pander # Exponentiate to get ORs
```


